---
title: "interception records with diagnostic results"
author: "Dalon White"
date: "2025-06-13"
output: html_document

params:
  CATEGORY: ["AQI"]
  SUBCATEGORY: ["CBP","PIS"]
  begin_date: 2020-01-01
  end_date: 
  commodities: ['Philodendron','philo','citrus']
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r load packages}
install.packages(pacman)
pacman::p_load(tidyverse, dplyr, stringr)
```

# Identify categories/subcategories

Call the category and subcategories in the yaml header

Categories can be "AQI" or "predeparture" or "domestic"

Subcategories can be "PIS", "CBP", "SITC", "Non-PIS PPQ"

```{r create environmental variables from the yaml header}

cats = params$CATEGORY
subcats = params$SUBCATEGORY

begin_date <- as.Date(params$begin_date)
#End date is the specified end date or the current date
end_date <- as.Date(if(exists("params$end_date")==FALSE){
  Sys.Date()
} else {
  params$end_date
})

commodities = params$commodities
```

## Retrieve pest identifications

```{r query-pest-diagnostics}
#function for connecting to database####
connect_db <- function(database) {
  DBI::dbConnect(
                 odbc::odbc(),
                 .connection_string =
"Driver=SQL Server;
                       Server=AAP00VA3PPQSQL0\\MSSQLSERVER,1433;
                        Database=PPQ_AQI_ARMDMV2;
trusted_connection=yes")
}
#connection to database
db_conn <- connect_db(database=db)

#We're going to use results because _Current doesn't ahve inspection date or category. Noah said this table includes tentative IDs as well but we'll filter those out anyways so that doesn't matter
mvw_diagnostic_results <- tbl(
  db_conn,
  sql("select TOP 10000 * FROM [APHIS_Imports].[dbo].[mvw_Diagnostic_Results]")
)

df_diagnostic_results <- mvw_diagnostic_results %>% 
  filter(
    INSPECTION_DATE >= begin_date,
    INSPECTION_DATE <= end_date,
    DETERMINATION_TYPE == 'Final ID',
    CATEGORY %in% cats,
    SUBCATEGORY %in% subcats,
    !is.na(PEST_TAXONOMIC_NAME)
  ) %>% 
  collect()
```

## Change quarantine status to be quarantine status at the POE

This is an optional chunk; the code will run if you skip it

Basically this will identify where the commodity came through (PoE, not PoArrival) and determine the quarantine status at that location.

```{r correct quarantine status to be quarantine status at POE}
#call diagnostic determination data; written flexibly because QUARANTINE_STATUS_MAINLAND just chagned to QUARANTINE_STATUS_CONUS and it differs by table
get.determinations <- function(connection) {
  # First, get all columns to avoid SQL errors if column names change
  result <- tbl(connection, sql("SELECT TOP 10000
                       *
                       FROM [PPQ_AQI_ARMDMV2].[ARMDATADM].[SYS2_BRG_DIAGNOSTIC_DETERMINATION]")) |> 
    collect()
  
  # Get the actual column names from the retrieved data
  actual_cols <- colnames(result)
  
  # Define mandatory columns we need
  base_cols <- c(
    "ID",
    "DIAGNOSTIC_REQUEST_ID",
    "PEST_TAXON_ID",
    "PEST_TAXON_SIMPLE_NAME",
    "PEST_TAXONOMIC_NAME",
    "ID_AUTHORITY",
    "DETERMINATION_TYPE_ID",
    "DETERMINED_BY_GROUP_ID",
    "DETERMINATION_DATETIME",
    "QUARANTINE_STATUS_HAWAII",
    "QUARANTINE_STATUS_PUERTO_RICO"
  )
  
  # Check which quarantine status columns exist and add them
  quarantine_mainland_col <- NULL
  if ("QUARANTINE_STATUS_MAINLAND" %in% actual_cols) {
    quarantine_mainland_col <- "QUARANTINE_STATUS_MAINLAND"
  } else if ("QUARANTINE_STATUS_CONUS" %in% actual_cols) {
    quarantine_mainland_col <- "QUARANTINE_STATUS_CONUS"
  }
  
  # Create the final column selection list
  select_cols <- c(base_cols, quarantine_mainland_col)
  select_cols <- select_cols[!is.null(select_cols)]
  
  # Select the needed columns
  result <- result |> 
    dplyr::select(all_of(select_cols)) |> 
    rename("DIAGNOSTIC_DETERMINATION_ID" = "ID")
  
  # If needed, standardize column names for downstream code
  if (!is.null(quarantine_mainland_col) && quarantine_mainland_col == "QUARANTINE_STATUS_MAINLAND") {
    result <- result |> rename("QUARANTINE_STATUS_CONUS" = "QUARANTINE_STATUS_MAINLAND")
  }
  
  return(result)
}

```

# Filter diagnostic determinations to the most authoritative final ID

unfortunately the dataset drops the diagnostic group number which is the easiest way to identify hierarchy in who's the most authoritative. So, connect that back first

```{r connect diagnostician expertise}
get_determination_group <- function(connection) {
   result = tbl(
  db_conn,
  sql("select TOP 10000 * FROM [PPQ_AQI_ARMDMV2].[ARMDATADM].[REF_DETERMINED_BY_GROUP]")
   ) |> 
     dplyr::select(ID,DETERMINED_GROUP) |> 
     rename('DETERMINED_BY_GROUP_ID' = 'ID') |> 
     collect()
}

determination_group <- get_determination_group(db_conn)

df_diagnostic_results <- df_diagnostic_results |> left_join(determination_group, by = c("DETERMINED_BY_GROUP_NAME" = "DETERMINED_GROUP"))
```

It also does not keep track of the ID type number

```{r connect ID type}
get_determination_type <- function(connection) {
   result = tbl(
  db_conn,
  sql("select TOP 10000 * FROM [PPQ_AQI_ARMDMV2].[ARMDATADM].[REF_DETERMINATION_TYPE]")
   ) |> 
     dplyr::select(ID,DETERMINATION_TYPE) |> 
     rename('DETERMINATION_TYPE_ID' = 'ID') |> 
     collect()
}

determination_type <- get_determination_type(db_conn)

df_diagnostic_results <- df_diagnostic_results |> left_join(determination_type, by = c("DETERMINATION_TYPE" = "DETERMINATION_TYPE"))
```

```{r parse diagnostic determinations}
df_diagnostic_results = if(exists("diag.determ.records") == TRUE){
  df_diagnostic_results |> 
    left_join(diag.determ.records,by=intersect(colnames(df_diagnostic_results), colnames(diag.determ.records)
                                               )
    )
  } else {
    df_diagnostic_results 
         }

parse_final_determination = function(data) {
  data |> 
    group_by(DIAGNOSTIC_REQUEST_ID) |> 
    #Andy Carmichael says if that the most recent "DETERMINATION_CREATED_DATETIME" should be the most accurate, with very rare exceptions.  (The only time I can think of is if there was a molecular ID request, and it failed.)   In that case there DETERM_NOT_POSS_FLAG should be flagged.  
    #Filter for rows that are not a failed diagnostic determination
    dplyr::filter(is.na(DIAGNOSTIC_DETERMINATION_NOT_POSSIBLE_REASON)) |> 
    #Filter the determination type - final has the highest TYPE_ID number
    dplyr::filter(DETERMINATION_TYPE_ID==max(DETERMINATION_TYPE_ID)) |> 
    #Filter for who ID'd it - the group_ID is a hierarchy of expertise, so take the max ID
    dplyr::filter(DETERMINED_BY_GROUP_ID == max(DETERMINED_BY_GROUP_ID)) |> 
    #The most recent is in practice the most correct - sometimes a final ID will be corrected at a later date, or someone will review for training or spot checking
    dplyr::filter(DETERMINATION_DATETIME == max(DETERMINATION_DATETIME))
}

df_diagnostic_results <- parse_final_determination(df_diagnostic_results)


```

# Retrieve commodity ID

If you are using mvw_diagnostic_results it will have commodity info

```{r sort by commodity}
df_diagnostic_results |> 
  mutate(COMMODITY_TAXONOMIC_DISPLAY_NAME = tolower(COMMODITY_TAXONOMIC_DISPLAY_NAME)) |> 
  filter(str_detect(
    COMMODITY_TAXONOMIC_DISPLAY_NAME,
    stringr::str_c(commodities,collapse = "|")
    )
    )
```

If you are using mvw_diagnostic_results_Current it will not have commodity info and you will have to conenct them yourself

{I haven't finished this as of 6/13/25}

```{r retrieving commodity info}
#### Pulling all commodity data, then filtering from there ----
#Use that to select for commodities
get.commodity = function(connection,
                         commodity
                         ) {
  tbl(connection, sql( "SELECT TOP 10000
                       *
                       FROM [PPQ_AQI_ARMDMV2].[ARMDATADM].[SYS2_BRG_COMMODITY]")) |> 
    filter(COMMODITY_COMMON_NAME %in% commodity |
             COMMODITY_DISPLAY_NAME %in% commodity) |>
    dplyr::select(
      ID,
      INSPECTION_ID,
      DIAGNOSTIC_EVENT_ID,
      QUANTITY,
      QUANTITY_UNITS_NAME,
      COMMODITY_HOST_TYPE,
      DISPOSITION_CODE,
      CBP_COMMODITY_ID_NUMBER,
      INITIAL_COMMODITY_DISPLAY_NAME,
      INITIAL_COMMODITY_TAXONOMIC_DISPLAY_NAME,
      INITIAL_COMMODITY_COMMON_NAME,
      COMMODITY_DISPLAY_NAME,
      COMMODITY_TAXONOMIC_DISPLAY_NAME,
      COMMODITY_COMMON_NAME,
      TAXON_SIMPLE_NAME,
      PROPAGATIVE_MATERIAL_TYPE,
      FINAL_TAXON_SIMPLE_NAME,
      DESTINATION_CITY,
      SHIPMENT_IDENTIFIER_ID,
      DIAGNOSTIC_EVENT_ID,
      COUNTRY_OF_ORIGIN_ID, #these aren't complete records
      COUNTRY_OF_ORIGIN_NAME, #these aren't complete records
      QUARANTINE_RECOMMENDATION
    ) |> 
    collect() |> 
    rename('COMMODITY_ID' = 'ID')
}

commodities = get.commodity(connection = db_conn)
```

# Connect diagnostic info to commodity ID

```{r connect pest determination to commodity info}

df_diagnostic_results |> left_join(commodities, by=c('COMMODITY_ID'))

```
