---
title: "OPEP FNW Select Program Risks"
author: "Dalon White"
date: "2025-05-08"
output: html_document

params:
  quarantine.pest.value: 0.5
  uncategorized.pest.value: 0.25
  nonquarantine.pest.value: 0
  caps.national.priority.pest.score: 1
  opep.noncandidate.pest.score: .3  # Based on OPEP list cutoff of 10% major invader
  ppq.program.pest.score: 1
  ppq.select.agents.score: 1
  begin_date: 2019-10-01
  end_date: 2025-04-01

---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r load packages, include=FALSE}
require(pacman)
pacman::p_load(
  tidyverse,
  magrittr,
  ggplot2,
  DBI,
  DT,
  odbc,
  lubridate,
  readr
  )
```

```{r get vars for global env from yaml header}
quarantine.pest.value <- params$quarantine.pest.value
uncategorized.pest.value <- params$uncategorized.pest.value
nonquarantine.pest.value <- params$nonquarantine.pest.value
caps.national.priority.pest.score <- params$caps.national.priority.pest.score
opep.noncandidate.pest.score <- params$opep.noncandidate.pest.score # Based on OPEP list cutoff of 10% major invader
ppq.program.pest.score <- params$ppq.program.pest.score
ppq.select.agents.score <- params$ppq.select.agents.score

begin_date <- params$begin_date
end_date <- params$end_date
```
## Known high risk pests

put together PPQ's known risks that extend beyond quarantine/non quarantine
Intent is for chunks to be incorporated into port profiles as appropriate

```{r load-and-prep-opep-data}
##THE OPEP LIST IS NATIVELY FORMATTED WITH EXTRA LINES AND STUFF - IT HAS TO BE MANUALLY MANIPULATED TO FIT A STANDARD DATA FORMAT AND SAVED .CSV
#Cynthia Landry stated that all the arthropod pests on the OPEP list will have genus/species names
#Cynthia has an alterantive list she created that uses the OPEP name they use when completing it; she thought this might be more in line with the NIS names for all the pathogens, so this is the one we use. Cynthia is the owner of that and this file should be updated probably every 3 months but at least yearly

files.in.opep = file.info(list.files(here::here('Inputs','Data','OPEP'), full.names = T))
##gets the file with the last modified time
opep.list = read.csv(rownames(files.in.opep)[which.max(files.in.opep$mtime)])
#get rid of some of the characters encoded in the names
opep.list$Scientific.name = iconv(opep.list$Scientific.name, "latin1", "ASCII", sub="")
opep.list$Common.Name = iconv(opep.list$Common.Name, "latin1", "ASCII", sub="")
opep.list = opep.list |>
  mutate(Scientific.name=trimws(Scientific.name),
         Common.Name=trimws(Common.Name),
         Genus = NA,
         Species = NA)

opep.list = opep.list |> 
  #extract genus name
  mutate(Genus = case_when(
    Model == 'Arthropod' ~ stringr::word(Scientific.name, 1),
    Model == 'Pathogen' ~ NA,
    .default= NA
  )) |> 
  #extract species name
  mutate(Species = case_when(
    Model == 'Arthropod' ~ stringr::word(Scientific.name, 2),
    Model == 'Pathogen' ~ NA,
    .default= NA
  ))

#arthropod genus and species (got to get rid of the NA's or else they will match in the next step)
arth.genus.species = opep.list |> filter(Model=='Arthropod') |> 
  filter(!Genus == '' & !Species=='') |> 
  dplyr::select(Genus, Species)

#scientific and common pathogen names
path.sci.name = opep.list |> filter(Model=='Pathogen') |> 
  pull(Scientific.name)
path.comm.name = opep.list |> filter(Model=='Pathogen') |> 
  filter(!Common.Name=='') |> 
  pull(Common.Name)

pest_impact_calc_func=function(high.impacts,mod.impacts){
  (
    (high.impacts*2) + mod.impacts) / 200
}

# determine quantification of impact for each OPEP pest and determine if the pest is in OPEP list ----
opep.list.red = opep.list |>
  #Remove lines that don't correspond to an organism
  filter(!Scientific.name=='') |>
  #If we don't have better information then just delete from list and it will default to a quarantine/non quarantine status
  filter(!Predicted.pest.impact.in.US %in% c('N/A','Undetermined')) |>
  mutate(Prob.pest.will.cause.high.impacts = parse_number(Prob.pest.will.cause.high.impacts),
         Prob.pest.will.cause.mod.impacts = parse_number(Prob.pest.will.cause.mod.impacts),
         Prob.pest.will.cause.low.impacts = parse_number(Prob.pest.will.cause.low.impacts)
  ) |> 
  #recalculate opep ranking
  mutate(Predicted.pest.impact.in.US = pest_impact_calc_func(Prob.pest.will.cause.high.impacts,
                                                             Prob.pest.will.cause.mod.impacts)
  #        ) |> 
  # #change name structure to match other data base
  # dplyr::select(Scientific.name,Common.Name,Genus,Species,Predicted.pest.impact.in.US) |> 
  # mutate(name.match=paste(Genus,Species,sep = ' ')) |> dplyr::select(-Genus,-Species)
)


#In the original version, I compare these lists to taxonomies in the inspection+commodity+diagnostic request (complete.records)
# see ~line 661
```
Now need to see where `opep.list.red` fits in port profiles



```{r load-FNW-data}
FNW.files.in.outputs = file.info(list.files(here::here('Inputs','Data','FNW'), full.names = T))
##gets the file with the last modified time.. this should work well in a pipeline system
FNW.dat <- read_csv(
  rownames(
    #only look for .csv files
    FNW.files.in.outputs |> filter(grepl('.csv',rownames(FNW.files.in.outputs)))
    #now find the most recently modified
    )[which.max(FNW.files.in.outputs$mtime)])

#Taxonomies are coming in as 0's instead of NAs. Some columns are all zeroes so reading as numerical.
FNW.dat <- FNW.dat |>
  mutate(across(starts_with("Taxonomy"), as.character)) |> 
  mutate(across(starts_with("Taxonomy"), ~na_if(.,"0")))

# Quantify risk -----

#fix 'N/A's
FNW.dat$`% Suitable Reported` = replace(FNW.dat$`% Suitable Reported`,FNW.dat$`% Suitable Reported`=='N/A',NA)
FNW.dat$`% Suitable Reported` = as.numeric(FNW.dat$`% Suitable Reported`)
FNW.dat = FNW.dat |> mutate(FNW_score =
                    #risk score based on likelihood its a major invader
                    (`% Major Invader`
                  +
                    #include minor invader to move risk away from 0 faster
                    (0.5*`% Minor Invader`))
                  ) |> 
  #adjust to a 0 - 1 scale
  mutate(FNW_score = FNW_score/100) |> 
  #modified by how much suitable area it has already covered
  mutate(FNW_score = case_when(
    # if % suitable area is NA then don't do anything
    is.na(`% Suitable Reported`) ~ FNW_score,
    # else if it is a number then adjust the risk by the amount of habitable space it already covers
    .default = FNW_score*(1-(`% Suitable Reported`/100))
  ))

# #Looking at the cutoffs for min and max values, it looks like 'Low' is % major invader ~0-10%, Mod is % Major Invader ~11-50%, and High is %Major Invader ~ >50%
# FNW.dat |> dplyr::select(`Name in ARM`,`WRA Risk Result`,`% Major Invader`,`% Minor Invader`) |> filter(!is.na(`WRA Risk Result`)) |> group_by(`WRA Risk Result`) |> 
#   slice_max(`% Major Invader`,n=1)
# FNW.dat |> dplyr::select(`Name in ARM`,`WRA Risk Result`,`% Major Invader`,`% Minor Invader`) |> filter(!is.na(`WRA Risk Result`)) |> group_by(`WRA Risk Result`) |>
#   slice_min(`% Major Invader`,n=1)
### Nevermind, ignore the above - see next line
#Per conversation (12/4/24) with Leah, who consulted her notes, Tony suggested assigning risk  similarly to other quarantine pests regardless of the qualitative rating assigned

#if no WRA assessment, assign it a quarantine rank value
FNW.dat = FNW.dat |> mutate(FNW_score = case_when(
  is.na(`% Major Invader`) ~ quarantine.pest.value,
  .default = FNW_score
))


# Merge with records -----
FNW.split = split(FNW.dat,
                  FNW.dat$`Taxon Unit Type`)

## Below is how I joined the risk data to actual rows of inspection/pest interceptions. This will probably have to happen with diagnostic results table, now.

# 
# #The following connected it by taxonomy 
# FNW_joined_list <- lapply(FNW.split, function(subset) {
#   if (subset$`Taxon Unit Type`[1] == "Genus") {
#     # For genus only, match genus and family
#     merge(complete.records, subset,  by.x=c('TAXONOMY_FAMILY','TAXONOMY_GENUS'),
#           #keep the y in each
#           by.y=c('Taxonomy Family','Taxonomy Genus'), all.y = TRUE)
#     #determine which group it is by identifying the first entry in the identify.by column within the subset
#   } else if (subset$`Taxon Unit Type`[1] == "Species") {
#     # For species-level, identify by genus and smpecies
#     merge(complete.records, subset, by.x=c('TAXONOMY_GENUS','TAXONOMY_SPECIES'),
#           by.y=c('Taxonomy Genus','Taxonomy Species'), all.y = TRUE)
#   } else if (subset$`Taxon Unit Type`[1] == "Subspecies") {
#     # For subspecies-level, identify by genus, species, and subspecies
#     merge(complete.records, subset, by.x=c('TAXONOMY_GENUS','TAXONOMY_SPECIES','TAXONOMY_SUBSPECIES'),
#           by.y=c('Taxonomy Genus','Taxonomy Species','Taxonomy Subspecies'), all.y = TRUE)
#   } else {
#     # the last group is variety but these are only identified to genus/species in the FNW list, so just keep the genus/species combination
#     merge(complete.records, subset, by.x=c('TAXONOMY_GENUS','TAXONOMY_SPECIES'),
#           by.y=c('Taxonomy Genus','Taxonomy Species'), all.y=TRUE)
#   }
# })
```
## Include program pest list

```{r load-HLI-data}
HLI.files.in.outputs = file.info(list.files(here::here('Inputs','Data','High and low impact pests'), full.names = T))
##gets the file with the last modified time.. this should work well in a pipeline system
HLI.dat = read_csv(
  rownames(
    #only look for .csv files
    HLI.files.in.outputs |> filter(grepl('.csv',rownames(HLI.files.in.outputs)))
    #now find the most recently modified
  )[which.max(HLI.files.in.outputs$mtime)])
HLI.dat = HLI.dat |> mutate(identify.by = case_when(
  `Pest Type` %in% 'Plant Disease' ~ 'pathogen name',
  str_detect(`Pest Scientific Name`,'. .') ~ 'genus and species',
  .default = 'genus only'))

#Quantify the risk score
HLI.dat = HLI.dat |> mutate(HLI_score = case_when(
  grepl('CAPS National Priority',Source) == T ~ caps.national.priority.pest.score,
  grepl('OPEP Non-Candidate',Source) == T ~ opep.noncandidate.pest.score,
  grepl('PPQ Program Pest',Source) == T ~ ppq.program.pest.score,
  grepl('Select Agents',Source) == T ~ ppq.select.agents.score,
  .default=0
))

HLI.split=split(HLI.dat,
  HLI.dat$identify.by)


## Below is how I joined the risk data to actual rows of inspection/pest interceptions. This will probably have to happen with diagnostic results table, now.


# # Join each subset to A, customizing the join type for each group
# HLI_joined_list <- lapply(HLI.split, function(subset) {
#   if (subset$identify.by[1] == "pathogen name") {
#     # After looking through the data, it appears the genus and species in both data sets would match
#     merge(complete.records, subset,  by.x=c('TAXONOMY_GENUS','TAXONOMY_SPECIES'),
#           #keep the y in each
#           by.y=c('Genus','Species'), all.y = TRUE)
#     #determine which group it is by identifying the first entry in the identify.by column within the subset
#   } else if (subset$identify.by[1] == "genus only") {
#     # For this group, only the genus is identified in the high/low impact dataset.
#     # Looking through the ARM records, when genus is identified, family is also identified
#     #   see 
#        #view(complete.records |> filter(!is.na(TAXONOMY_SPECIES)) |> count(TAXONOMY_FAMILY))
#     merge(complete.records, subset, by.x=c('TAXONOMY_FAMILY','TAXONOMY_GENUS'),
#           by.y=c('Family','Genus'), all.y = TRUE)
#   } else {
#     # the only other group is genus and species, so default to comparing genus and species
#     merge(complete.records, subset, by.x=c('TAXONOMY_GENUS','TAXONOMY_SPECIES'),
#           by.y=c('Genus','Species'), all.y=TRUE)
#   }
# })
# 
# # Combine the joined dataframes
# # use dplyr::bind_rows instead of rbind because the matching columns rearrange the column order;
# #   bind_rows will match column names instead of requiring the same order
# HLI_final_df <- do.call(bind_rows, HLI_joined_list)
# 
# # #does it join correctly?
# # complete.records |> left_join(final_df)
# # #is it matching genus and species correctly? do they have the same name?
# #  complete.records |> left_join(final_df) |> filter(!is.na(`Qualitative Impact Rating`)) |>
# #    dplyr::select(TAXONOMY_GENUS,TAXONOMY_SPECIES,`Pest Scientific Name`)
# # #has it duplicated anything?
# # which(duplicated(complete.records) == TRUE) #this code works correctly, as evidenced by changing TRUE to FALSE
# # #everything appears to be working correctly
# 
# complete.records = complete.records |> left_join(HLI_final_df) |> dplyr::select(
#   #all of complete.records columns
#   colnames(complete.records),
#   #source of info, pest impact, and scientific name for completeness
#   Source,
#   `Qualitative Impact Rating`,
#   `Pest Scientific Name`,
#   HLI_score)

```


```{r query-pest-diagnostics}
#function for connecting to database####
connect_db <- function(database) {
  DBI::dbConnect(
                 odbc::odbc(),
                 .connection_string =
"Driver=SQL Server;
                       Server=AAP00VA3PPQSQL0\\MSSQLSERVER,1433;
                        Database=PPQ_AQI_ARMDMV2;
trusted_connection=yes")
}
#connection to database
dbconn <- connect_db(database=db)

mvw_diagnostic_results <- tbl(
  db_conn,
  sql("select TOP 1000 * FROM [APHIS_Imports].[dbo].[mvw_Diagnostic_Results]")
)

df_diagnostic_results <- mvw_diagnostic_results %>% 
  filter(
    INSPECTION_DATE >= begin_date,
    INSPECTION_DATE <= end_date,
    DETERMINATION_TYPE == 'Final ID',
    SUBCATEGORY %in% c("CBP", "PIS", "SITC", "Non-PIS PPQ"),
    !is.na(PEST_TAXONOMIC_NAME)
  ) %>% 
  collect()
```

```{r correct query_pest_diagnostics}
#Final ID are repeated, at least because of different life stages of the same pest


# #replaced the below function with a flexible one
# get.determinations = function(connection,determ.ids) {
#   tbl(connection, sql( "SELECT TOP 100000
#                        *
#                        FROM [PPQ_AQI_ARMDMV2].[ARMDATADM].[SYS2_BRG_DIAGNOSTIC_DETERMINATION]")) |> 
#     #    filter(ID %in% determ.ids) |> 
#     dplyr::select(
#       ID,
#       DIAGNOSTIC_REQUEST_ID,
#       PEST_TAXON_ID,
#       PEST_TAXON_SIMPLE_NAME,
#       PEST_TAXONOMIC_NAME,
#       QUARANTINE_STATUS_CONUS,
#       QUARANTINE_STATUS_HAWAII,
#       QUARANTINE_STATUS_PUERTO_RICO,
#       ID_AUTHORITY,
#       DETERMINATION_TYPE_ID,
#       DETERMINED_BY_GROUP_ID,
#       DETERMINATION_DATETIME
#     ) |> rename("DIAGNOSTIC_DETERMINATION_ID" = "ID") |> 
#     collect()
# }


#call diagnostic determination data; written flexibly because QUARANTINE_STATUS_MAINLAND just chagned to QUARANTINE_STATUS_CONUS and it differs by table
get.determinations <- function(connection) {
  # First, get all columns to avoid SQL errors if column names change
  result <- tbl(connection, sql("SELECT TOP 100000
                       *
                       FROM [PPQ_AQI_ARMDMV2].[ARMDATADM].[SYS2_BRG_DIAGNOSTIC_DETERMINATION]")) |> 
    collect()
  
  # Get the actual column names from the retrieved data
  actual_cols <- colnames(result)
  
  # Define mandatory columns we need
  base_cols <- c(
    "ID",
    "DIAGNOSTIC_REQUEST_ID",
    "PEST_TAXON_ID",
    "PEST_TAXON_SIMPLE_NAME",
    "PEST_TAXONOMIC_NAME",
    "ID_AUTHORITY",
    "DETERMINATION_TYPE_ID",
    "DETERMINED_BY_GROUP_ID",
    "DETERMINATION_DATETIME",
    "QUARANTINE_STATUS_HAWAII",
    "QUARANTINE_STATUS_PUERTO_RICO"
  )
  
  # Check which quarantine status columns exist and add them
  quarantine_mainland_col <- NULL
  if ("QUARANTINE_STATUS_MAINLAND" %in% actual_cols) {
    quarantine_mainland_col <- "QUARANTINE_STATUS_MAINLAND"
  } else if ("QUARANTINE_STATUS_CONUS" %in% actual_cols) {
    quarantine_mainland_col <- "QUARANTINE_STATUS_CONUS"
  }
  
  # Create the final column selection list
  select_cols <- c(base_cols, quarantine_mainland_col)
  select_cols <- select_cols[!is.null(select_cols)]
  
  # Select the needed columns
  result <- result |> 
    dplyr::select(all_of(select_cols)) |> 
    rename("DIAGNOSTIC_DETERMINATION_ID" = "ID")
  
  # If needed, standardize column names for downstream code
  if (!is.null(quarantine_mainland_col) && quarantine_mainland_col == "QUARANTINE_STATUS_MAINLAND") {
    result <- result |> rename("QUARANTINE_STATUS_CONUS" = "QUARANTINE_STATUS_MAINLAND")
  }
  
  return(result)
}

diag.determ.records = get.determinations(connection=db_conn) |> 
  filter(DIAGNOSTIC_DETERMINATION_ID %in% (df_diagnostic_results |> pull(DIAGNOSTIC_DETERMINATION_ID)))

df_diagnostic_results = df_diagnostic_results |> left_join(diag.determ.records,by=intersect(colnames(df_diagnostic_results),colnames(diag.determ.records))) 

parse_final_determination = function(data) {
  data |> 
    group_by(DIAGNOSTIC_REQUEST_ID) |> 
    #Andy Carmichael says if that the most recent "DETERMINATION_CREATED_DATETIME" should be the most accurate, with very rare exceptions.  (The only time I can think of is if there was a molecular ID request, and it failed.)   In that case there DETERM_NOT_POSS_FLAG should be flagged.  
    #Filter for rows that are not a failed diagnostic determination
    dplyr::filter(is.na(DIAGNOSTIC_DETERMINATION_NOT_POSSIBLE_REASON)) |> 
    #Filter the determination type - final has the highest TYPE_ID number
    dplyr::filter(DETERMINATION_TYPE_ID==max(DETERMINATION_TYPE_ID)) |> 
    #Filter for who ID'd it - the group_ID is a hierarchy of expertise, so take the max ID
    dplyr::filter(DETERMINED_BY_GROUP_ID == max(DETERMINED_BY_GROUP_ID)) |> 
    #The most recent is in practice the most correct - sometimes a final ID will be corrected at a later date, or someone will review for training or spot checking
    dplyr::filter(DETERMINATION_DATETIME == max(DETERMINATION_DATETIME))
}

df_diagnostic_results <- parse_final_determination(df_diagnostic_results)

#Left off here 5/8.
#The above code is working. it's flexible and solves for MAINLAND vs CONUS
#it also integrates with the DART's diagnostic results table but parses using our approach as well.

```
Next step is to assess the df_diagnostic_results taxonomies for matches across FNW, HLI, or OPEP

```{r assign_pest_score_to_each_record}
#Check later, but I think the quarantine status might be coming from CONUS; use my code to assign quarantine status based on port location

#df_diagnostic_results

# Join each subset to A, customizing the join type for each group
HLI_joined_list <- lapply(HLI.split, function(subset) {
  if (subset$identify.by[1] == "pathogen name") {
    # After looking through the data, it appears the genus and species in both data sets would match
    merge(df_diagnostic_results, subset,  by.x=c('TAXONOMY_GENUS','TAXONOMY_SPECIES'),
          #keep the y in each
          by.y=c('Genus','Species'), all.y = TRUE)
    #determine which group it is by identifying the first entry in the identify.by column within the subset
  } else if (subset$identify.by[1] == "genus only") {
    # For this group, only the genus is identified in the high/low impact dataset.
    # Looking through the ARM records, when genus is identified, family is also identified
    #   see
       #view(df_diagnostic_results |> filter(!is.na(TAXONOMY_SPECIES)) |> count(TAXONOMY_FAMILY))
    merge(df_diagnostic_results, subset, by.x=c('TAXONOMY_FAMILY','TAXONOMY_GENUS'),
          by.y=c('Family','Genus'), all.y = TRUE)
  } else {
    # the only other group is genus and species, so default to comparing genus and species
    merge(df_diagnostic_results, subset, by.x=c('TAXONOMY_GENUS','TAXONOMY_SPECIES'),
          by.y=c('Genus','Species'), all.y=TRUE)
  }
})

# Combine the joined dataframes
# use dplyr::bind_rows instead of rbind because the matching columns rearrange the column order;
#   bind_rows will match column names instead of requiring the same order
HLI_final_df <- do.call(bind_rows, HLI_joined_list)

# #does it join correctly?
# df_diagnostic_results |> left_join(final_df)
# #is it matching genus and species correctly? do they have the same name?
#  df_diagnostic_results |> left_join(final_df) |> filter(!is.na(`Qualitative Impact Rating`)) |>
#    dplyr::select(TAXONOMY_GENUS,TAXONOMY_SPECIES,`Pest Scientific Name`)
# #has it duplicated anything?
# which(duplicated(df_diagnostic_results) == TRUE) #this code works correctly, as evidenced by changing TRUE to FALSE
# #everything appears to be working correctly

df_diagnostic_results = df_diagnostic_results |> left_join(HLI_final_df) |> dplyr::select(
  #all of complete.records columns
  colnames(df_diagnostic_results),
  #source of info, pest impact, and scientific name for completeness
  Source,
  `Qualitative Impact Rating`,
  `Pest Scientific Name`,
  HLI_score)



```

Now find out where the port profile gets commodity information.


