---
title: "OPEP FNW Select Program Risks"
author: "Dalon White"
date: "2025-05-08"
output: html_document

params:
  quarantine.pest.value: 0.5
  uncategorized.pest.value: 0.25
  nonquarantine.pest.value: 0
  caps.national.priority.pest.score: 1
  opep.noncandidate.pest.score: .3  # Based on OPEP list cutoff of 10% major invader
  ppq.program.pest.score: 1
  ppq.select.agents.score: 1
  begin_date: 2019-10-01
  end_date: 2025-04-01
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r load packages, include=FALSE}
require(pacman)
pacman::p_load(
  tidyverse,
  magrittr,
  ggplot2,
  DBI,
  DT,
  odbc,
  lubridate,
  readr
  )
```

```{r get vars for global env from yaml header}
quarantine.pest.value <- params$quarantine.pest.value
uncategorized.pest.value <- params$uncategorized.pest.value
nonquarantine.pest.value <- params$nonquarantine.pest.value
caps.national.priority.pest.score <- params$caps.national.priority.pest.score
opep.noncandidate.pest.score <- params$opep.noncandidate.pest.score # Based on OPEP list cutoff of 10% major invader
ppq.program.pest.score <- params$ppq.program.pest.score
ppq.select.agents.score <- params$ppq.select.agents.score

begin_date <- params$begin_date
end_date <- params$end_date
```

# Known pest impacts

Put together PPQ's known risks that extend beyond quarantine/non quarantine

Intent is for chunks to be incorporated into port profiles as appropriate

## OPEP list

Start with OPEP

```{r load-and-prep-opep-data}
##THE OPEP LIST IS NATIVELY FORMATTED WITH EXTRA LINES AND STUFF - IT HAS TO BE MANUALLY MANIPULATED TO FIT A STANDARD DATA FORMAT AND SAVED .CSV
#Cynthia Landry stated that all the arthropod pests on the OPEP list will have genus/species names
#Cynthia has an alterantive list she created that uses the OPEP name they use when completing it; she thought this might be more in line with the NIS names for all the pathogens, so this is the one we use. Cynthia is the owner of that and this file should be updated probably every 3 months but at least yearly

files.in.opep = file.info(list.files(here::here('Inputs','Data','OPEP'), full.names = T))
##gets the file with the last modified time
opep.list = read.csv(rownames(files.in.opep)[which.max(files.in.opep$mtime)])
#get rid of some of the characters encoded in the names
opep.list$Scientific.name = iconv(opep.list$Scientific.name, "latin1", "ASCII", sub="")
opep.list$Common.Name = iconv(opep.list$Common.Name, "latin1", "ASCII", sub="")
opep.list = opep.list |>
  mutate(Scientific.name=trimws(Scientific.name),
         Common.Name=trimws(Common.Name),
         Genus = NA,
         Species = NA)

opep.list = opep.list |> 
  #extract genus name
  mutate(Genus = case_when(
    Model == 'Arthropod' ~ stringr::word(Scientific.name, 1),
    Model == 'Pathogen' ~ NA,
    .default= NA
  )) |> 
  #extract species name
  mutate(Species = case_when(
    Model == 'Arthropod' ~ stringr::word(Scientific.name, 2),
    Model == 'Pathogen' ~ NA,
    .default= NA
  ))

#arthropod genus and species (got to get rid of the NA's or else they will match in the next step)
arth.genus.species = opep.list |> filter(Model=='Arthropod') |> 
  filter(!Genus == '' & !Species=='') |> 
  dplyr::select(Genus, Species)

#scientific and common pathogen names
path.sci.name = opep.list |> filter(Model=='Pathogen') |> 
  pull(Scientific.name)
path.comm.name = opep.list |> filter(Model=='Pathogen') |> 
  filter(!Common.Name=='') |> 
  pull(Common.Name)

#Create a function that calculates the expected impact of a pest based on high/mod/low impact probabilities (scaled 0-1)
pest_impact_calc_func=function(high.impacts,mod.impacts){
  (
    (high.impacts*2) + mod.impacts) / 200
}

# determine quantification of impact for each OPEP pest and determine if the pest is in OPEP list ----
opep.list.red = opep.list |> filter(!Scientific.name=='') |> 
  filter(!Predicted.pest.impact.in.US %in% c('N/A','Undetermined')) |> 
  # filter(is.na(Predicted.pest.impact.in.US)) |> 
  mutate(Prob.pest.will.cause.high.impacts = parse_number(Prob.pest.will.cause.high.impacts),
         Prob.pest.will.cause.mod.impacts = parse_number(Prob.pest.will.cause.mod.impacts),
         Prob.pest.will.cause.low.impacts = parse_number(Prob.pest.will.cause.low.impacts)
  ) |> 
  #recalculate opep ranking
  mutate(Predicted.pest.impact.in.US = ((Prob.pest.will.cause.high.impacts*2)+Prob.pest.will.cause.mod.impacts)/200) |> 
  #change name structure to match other data base
  dplyr::select(Scientific.name,Common.Name,Genus,Species,Predicted.pest.impact.in.US) |> 
  mutate(name.match=paste(Genus,Species,sep = ' ')) |> dplyr::select(-Genus,-Species)


# # determine quantification of impact for each OPEP pest and determine if the pest is in OPEP list ----
# opep.list.red = opep.list |>
#   #Remove lines that don't correspond to an organism
#   filter(!Scientific.name=='') |>
#   #If we don't have better information then just delete from list and it will default to a quarantine/non quarantine status
#   filter(!Predicted.pest.impact.in.US %in% c('N/A','Undetermined')) |>
#   mutate(Prob.pest.will.cause.high.impacts = parse_number(Prob.pest.will.cause.high.impacts),
#          Prob.pest.will.cause.mod.impacts = parse_number(Prob.pest.will.cause.mod.impacts),
#          Prob.pest.will.cause.low.impacts = parse_number(Prob.pest.will.cause.low.impacts)
#   ) |> 
#   #recalculate opep ranking
#   mutate(Predicted.pest.impact.in.US = pest_impact_calc_func(Prob.pest.will.cause.high.impacts,
#                                                              Prob.pest.will.cause.mod.impacts)
#   #        ) |> 
#   # #change name structure to match other data base
#   # dplyr::select(Scientific.name,Common.Name,Genus,Species,Predicted.pest.impact.in.US) |> 
#   # mutate(name.match=paste(Genus,Species,sep = ' ')) |> dplyr::select(-Genus,-Species)
# )


#In the original version, I compare these lists to taxonomies in the inspection+commodity+diagnostic request (complete.records)
# see ~line 661
```

## FNW list

```{r load-FNW-data, message = FALSE}
FNW.files.in.outputs = file.info(list.files(here::here('Inputs','Data','FNW'), full.names = T))
##gets the file with the last modified time.. this should work well in a pipeline system
FNW.dat <- read_csv(
  rownames(
    #only look for .csv files
    FNW.files.in.outputs |> filter(grepl('.csv',rownames(FNW.files.in.outputs)))
    #now find the most recently modified
    )[which.max(FNW.files.in.outputs$mtime)])

#Taxonomies are coming in as 0's instead of NAs. Some columns are all zeroes so reading as numerical.
FNW.dat <- FNW.dat |>
  mutate(across(starts_with("Taxonomy"), as.character)) |> 
  mutate(across(starts_with("Taxonomy"), ~na_if(.,"0")))

# Quantify risk -----

#fix 'N/A's
FNW.dat$`% Suitable Reported` = replace(FNW.dat$`% Suitable Reported`,FNW.dat$`% Suitable Reported`=='N/A',NA)
FNW.dat$`% Suitable Reported` = as.numeric(FNW.dat$`% Suitable Reported`)
FNW.dat = FNW.dat |> mutate(FNW_score =
                    #risk score based on likelihood its a major invader
                    (`% Major Invader`
                  +
                    #include minor invader to move risk away from 0 faster
                    (0.5*`% Minor Invader`))
                  ) |> 
  #adjust to a 0 - 1 scale
  mutate(FNW_score = FNW_score/100) |> 
  #modified by how much suitable area it has already covered
  mutate(FNW_score = case_when(
    # if % suitable area is NA then don't do anything
    is.na(`% Suitable Reported`) ~ FNW_score,
    # else if it is a number then adjust the risk by the amount of habitable space it already covers
    .default = FNW_score*(1-(`% Suitable Reported`/100))
  ))

# #Looking at the cutoffs for min and max values, it looks like 'Low' is % major invader ~0-10%, Mod is % Major Invader ~11-50%, and High is %Major Invader ~ >50%
# FNW.dat |> dplyr::select(`Name in ARM`,`WRA Risk Result`,`% Major Invader`,`% Minor Invader`) |> filter(!is.na(`WRA Risk Result`)) |> group_by(`WRA Risk Result`) |> 
#   slice_max(`% Major Invader`,n=1)
# FNW.dat |> dplyr::select(`Name in ARM`,`WRA Risk Result`,`% Major Invader`,`% Minor Invader`) |> filter(!is.na(`WRA Risk Result`)) |> group_by(`WRA Risk Result`) |>
#   slice_min(`% Major Invader`,n=1)
### Nevermind, ignore the above - see next line
#Per conversation (12/4/24) with Leah, who consulted her notes, Tony suggested assigning risk  similarly to other quarantine pests regardless of the qualitative rating assigned

#if no WRA assessment, assign it a quarantine rank value
FNW.dat = FNW.dat |> mutate(FNW_score = case_when(
  is.na(`% Major Invader`) ~ quarantine.pest.value,
  .default = FNW_score
))


# Merge with records -----
FNW.split = split(FNW.dat,
                  FNW.dat$`Taxon Unit Type`)

```

### FNW Delete me

When found where to incoproate FNW taxonomies

## Program pest list

```{r load-HLI-data, message = FALSE}
HLI.files.in.outputs = file.info(list.files(here::here('Inputs','Data','High and low impact pests'), full.names = T))
##gets the file with the last modified time.. this should work well in a pipeline system
HLI.dat = read_csv(
  rownames(
    #only look for .csv files
    HLI.files.in.outputs |> filter(grepl('.csv',rownames(HLI.files.in.outputs)))
    #now find the most recently modified
  )[which.max(HLI.files.in.outputs$mtime)])
HLI.dat = HLI.dat |> mutate(identify.by = case_when(
  `Pest Type` %in% 'Plant Disease' ~ 'pathogen name',
  str_detect(`Pest Scientific Name`,'. .') ~ 'genus and species',
  .default = 'genus only'))

#Quantify the risk score
HLI.dat = HLI.dat |> mutate(HLI_score = case_when(
  grepl('CAPS National Priority',Source) == T ~ caps.national.priority.pest.score,
  grepl('OPEP Non-Candidate',Source) == T ~ opep.noncandidate.pest.score,
  grepl('PPQ Program Pest',Source) == T ~ ppq.program.pest.score,
  grepl('Select Agents',Source) == T ~ ppq.select.agents.score,
  .default=0
))

HLI.split=split(HLI.dat,
  HLI.dat$identify.by)

```

### HLI delete me

When found where to incorporate HLI

### HLI Delete stale code

also delete stale code

# Retrieve pest query diagnostics

```{r query-pest-diagnostics}
#function for connecting to database####
connect_db <- function(database) {
  DBI::dbConnect(
                 odbc::odbc(),
                 .connection_string =
"Driver=SQL Server;
                       Server=AAP00VA3PPQSQL0\\MSSQLSERVER,1433;
                        Database=PPQ_AQI_ARMDMV2;
trusted_connection=yes")
}
#connection to database
db_conn <- connect_db(database=db)

mvw_diagnostic_results <- tbl(
  db_conn,
  sql("select TOP 10000 * FROM [APHIS_Imports].[dbo].[mvw_Diagnostic_Results]")
)

df_diagnostic_results <- mvw_diagnostic_results %>% 
  filter(
    INSPECTION_DATE >= begin_date,
    INSPECTION_DATE <= end_date,
    DETERMINATION_TYPE == 'Final ID',
    SUBCATEGORY %in% c("CBP", "PIS", "SITC", "Non-PIS PPQ"),
    !is.na(PEST_TAXONOMIC_NAME)
  ) %>% 
  collect()
```

## Correct pest query diagnostics

Correct based on quarantine status at port of entry as well as parse the final determinations to not double count multiple requests of the same pest because of different life stages or multiple identifiers.

```{r correct query_pest_diagnostics}
#Final ID are repeated, at least because of different life stages of the same pest


# #replaced the below function with a flexible one
# get.determinations = function(connection,determ.ids) {
#   tbl(connection, sql( "SELECT TOP 100000
#                        *
#                        FROM [PPQ_AQI_ARMDMV2].[ARMDATADM].[SYS2_BRG_DIAGNOSTIC_DETERMINATION]")) |> 
#     #    filter(ID %in% determ.ids) |> 
#     dplyr::select(
#       ID,
#       DIAGNOSTIC_REQUEST_ID,
#       PEST_TAXON_ID,
#       PEST_TAXON_SIMPLE_NAME,
#       PEST_TAXONOMIC_NAME,
#       QUARANTINE_STATUS_CONUS,
#       QUARANTINE_STATUS_HAWAII,
#       QUARANTINE_STATUS_PUERTO_RICO,
#       ID_AUTHORITY,
#       DETERMINATION_TYPE_ID,
#       DETERMINED_BY_GROUP_ID,
#       DETERMINATION_DATETIME
#     ) |> rename("DIAGNOSTIC_DETERMINATION_ID" = "ID") |> 
#     collect()
# }


#call diagnostic determination data; written flexibly because QUARANTINE_STATUS_MAINLAND just chagned to QUARANTINE_STATUS_CONUS and it differs by table
get.determinations <- function(connection) {
  # First, get all columns to avoid SQL errors if column names change
  result <- tbl(connection, sql("SELECT
                       *
                       FROM [PPQ_AQI_ARMDMV2].[ARMDATADM].[SYS2_BRG_DIAGNOSTIC_DETERMINATION]")) |> 
    collect()
  
  # Get the actual column names from the retrieved data
  actual_cols <- colnames(result)
  
  # Define mandatory columns we need
  base_cols <- c(
    "ID",
    "DIAGNOSTIC_REQUEST_ID",
    "PEST_TAXON_ID",
    "PEST_TAXON_SIMPLE_NAME",
    "PEST_TAXONOMIC_NAME",
    "ID_AUTHORITY",
    "DETERMINATION_TYPE_ID",
    "DETERMINED_BY_GROUP_ID",
    "DETERMINATION_DATETIME",
    "QUARANTINE_STATUS_HAWAII",
    "QUARANTINE_STATUS_PUERTO_RICO"
  )
  
  # Check which quarantine status columns exist and add them
  quarantine_mainland_col <- NULL
  if ("QUARANTINE_STATUS_MAINLAND" %in% actual_cols) {
    quarantine_mainland_col <- "QUARANTINE_STATUS_MAINLAND"
  } else if ("QUARANTINE_STATUS_CONUS" %in% actual_cols) {
    quarantine_mainland_col <- "QUARANTINE_STATUS_CONUS"
  }
  
  # Create the final column selection list
  select_cols <- c(base_cols, quarantine_mainland_col)
  select_cols <- select_cols[!is.null(select_cols)]
  
  # Select the needed columns
  result <- result |> 
    dplyr::select(all_of(select_cols)) |> 
    rename("DIAGNOSTIC_DETERMINATION_ID" = "ID")
  
  # If needed, standardize column names for downstream code
  if (!is.null(quarantine_mainland_col) && quarantine_mainland_col == "QUARANTINE_STATUS_MAINLAND") {
    result <- result |> rename("QUARANTINE_STATUS_CONUS" = "QUARANTINE_STATUS_MAINLAND")
  }
  
  return(result)
}

diag.determ.records = get.determinations(connection=db_conn) |> 
  filter(DIAGNOSTIC_DETERMINATION_ID %in% (df_diagnostic_results |> pull(DIAGNOSTIC_DETERMINATION_ID)))

df_diagnostic_results = df_diagnostic_results |> left_join(diag.determ.records,by=intersect(colnames(df_diagnostic_results),colnames(diag.determ.records))) 

parse_final_determination = function(data) {
  data |> 
    group_by(DIAGNOSTIC_REQUEST_ID) |> 
    #Andy Carmichael says if that the most recent "DETERMINATION_CREATED_DATETIME" should be the most accurate, with very rare exceptions.  (The only time I can think of is if there was a molecular ID request, and it failed.)   In that case there DETERM_NOT_POSS_FLAG should be flagged.  
    #Filter for rows that are not a failed diagnostic determination
    dplyr::filter(is.na(DIAGNOSTIC_DETERMINATION_NOT_POSSIBLE_REASON)) |> 
    #Filter the determination type - final has the highest TYPE_ID number
    dplyr::filter(DETERMINATION_TYPE_ID==max(DETERMINATION_TYPE_ID)) |> 
    #Filter for who ID'd it - the group_ID is a hierarchy of expertise, so take the max ID
    dplyr::filter(DETERMINED_BY_GROUP_ID == max(DETERMINED_BY_GROUP_ID)) |> 
    #The most recent is in practice the most correct - sometimes a final ID will be corrected at a later date, or someone will review for training or spot checking
    dplyr::filter(DETERMINATION_DATETIME == max(DETERMINATION_DATETIME))
}

df_diagnostic_results <- parse_final_determination(df_diagnostic_results)

#The above code is flexible and solves for MAINLAND vs CONUS
#it also integrates with the DART's diagnostic results table but parses using our approach as well.

```

# Integrate diagnostic results with known pest impacts

Next step is to assess the df_diagnostic_results taxonomies for matches across FNW, HLI, or OPEP

## Score pest taxonomies based on matches with OPEP list

```{r integrate_opep_impacts, message=FALSE}


# Prep a column in diagnostic records to accept OPEP pest info
df_diagnostic_results$OPEP.pest = grepl(
  paste(
    #Stitch back genus and species - separating was just done to get rid of white space and extra tags on the end of the genus/species
    paste(
      arth.genus.species$Genus,
      arth.genus.species$Species,
      sep=' ')
    ,collapse='|'),
  #find genus and species at the same time in order to avoid the situation where one genus is found and a species from another genus also matches
  paste(df_diagnostic_results$PEST_TAXONOMY_GENUS,df_diagnostic_results$PEST_TAXONOMY_SPECIES, sep=' ')
) #| grepl(
#   paste(trimws(path.sci.name),
#         trimws(path.comm.name),
#         collapse='|'),
#   df_diagnostic_results$PEST_TAXONOMIC_NAME
# )

#Match by pathogen scientific names. These are unlikely to be used in ARM
df_diagnostic_results$OPEP.path.sci.name = grepl(
  paste(trimws(path.sci.name),
        #         trimws(path.comm.name),
        collapse='|'),
  df_diagnostic_results$PEST_TAXONOMIC_NAME
)

#Match by pathogen common names. These are likely to be used to distinguish the pathogen in ARM
df_diagnostic_results$OPEP.path.comm.name = grepl(
  paste(#trimws(path.sci.name),
    trimws(path.comm.name),
    collapse='|'),
  df_diagnostic_results$PEST_TAXONOMIC_NAME
)

#The below is the 
df_diagnostic_results = df_diagnostic_results |> mutate(name.match=paste(PEST_TAXONOMY_GENUS,PEST_TAXONOMY_SPECIES,sep = ' ')) %>%
  left_join(opep.list.red |> filter(!name.match=='NA NA'), 
            by = c("name.match" = "name.match")) %>%
  left_join(opep.list.red, 
            by = c("PEST_TAXONOMIC_NAME" = "Scientific.name")) %>%
  mutate(Predicted.pest.impact.in.US_2 = coalesce(Predicted.pest.impact.in.US.x, Predicted.pest.impact.in.US.y)) %>%
  dplyr::select(-c(ends_with(".x"), ends_with(".y")))%>%
  left_join(opep.list.red, 
            by = c("PEST_TAXONOMIC_NAME" = "Common.Name")) %>%
  mutate(Predicted_pest_impact_in_US = coalesce(Predicted.pest.impact.in.US_2, Predicted.pest.impact.in.US)) %>%
  dplyr::select(-Predicted.pest.impact.in.US, -Predicted.pest.impact.in.US_2)


```

This could use a little cleaning up

-   check for extra columns

-   comb for stale code

## Score pest taxonomies based on matches with FNW list

```{r integrate_FNW_impacts, message = FALSE}

## Below is how I joined the risk data to actual rows of inspection/pest interceptions.

#The following connected it by taxonomy
FNW_joined_list <- lapply(FNW.split, function(subset) {
  if (subset$`Taxon Unit Type`[1] == "Genus") {
    # For genus only, match genus and family
    merge(df_diagnostic_results, subset,  by.x=c('PEST_TAXONOMY_FAMILY','PEST_TAXONOMY_GENUS'),
          #keep the y in each
          by.y=c('Taxonomy Family','Taxonomy Genus'), all.y = TRUE)
    #determine which group it is by identifying the first entry in the identify.by column within the subset
  } else if (subset$`Taxon Unit Type`[1] == "Species") {
    # For species-level, identify by genus and smpecies
    merge(df_diagnostic_results, subset, by.x=c('PEST_TAXONOMY_GENUS','PEST_TAXONOMY_SPECIES'),
          by.y=c('Taxonomy Genus','Taxonomy Species'), all.y = TRUE)
  } else if (subset$`Taxon Unit Type`[1] == "Subspecies") {
    # For subspecies-level, identify by genus, species, and subspecies
    merge(df_diagnostic_results, subset, by.x=c('PEST_TAXONOMY_GENUS','PEST_TAXONOMY_SPECIES'#,'TAXONOMY_SUBSPECIES'
                                                )
                                                ,
          by.y=c('Taxonomy Genus','Taxonomy Species'
                 #mvw_diagnostic_results does not have subspecies; additionally, the only subspecies identified currently in the FNW is Setaria pumila ssp. pallidefusca; according to NCSU extension page on the Setaria pumila, this subspecies no longer considered different from the species: "The root system if fibrous on this self-seeding specimen. Varieties and Subspecies, including one on the federal list of noxious weeds, are no longer recognized as distinct from the species. All plants of this species can have weedy and aggressive tendencies."
                 
                 #,'PEST_TAXONOMY_SUBSPECIES'
                 )
                 , all.y = TRUE)
  } else {
    # the last group is variety but these are only identified to genus/species in the FNW list, so just keep the genus/species combination
    merge(df_diagnostic_results, subset, by.x=c('PEST_TAXONOMY_GENUS','PEST_TAXONOMY_SPECIES'),
          by.y=c('Taxonomy Genus','Taxonomy Species'), all.y=TRUE)
  }
})

# Combine the joined dataframes
FNW_final_df <- do.call(bind_rows, FNW_joined_list) |> 
  rename(`Taxon_Identification_Level`=`Taxon Unit Type`)

df_diagnostic_results = df_diagnostic_results |> left_join(FNW_final_df) |> dplyr::select(
  #all of complete.records columns
  colnames(df_diagnostic_results),
  #source of info, pest impact, and scientific name for completeness
  Source,
  `Name in ARM`,
  `Taxon_Identification_Level`,
  FNW_score
  )
```

## Score pest taxonomies based on matches with HLI list

```{r assign_pest_score_to_each_record, message = FALSE}
#Check later, but I think the quarantine status might be coming from CONUS; use my code to assign quarantine status based on port location

#df_diagnostic_results

# Join each subset to A, customizing the join type for each group
HLI_joined_list <- lapply(HLI.split, function(subset) {
  if (subset$identify.by[1] == "pathogen name") {
    # After looking through the data, it appears the genus and species in both data sets would match
    merge(df_diagnostic_results, subset,  by.x=c('PEST_TAXONOMY_GENUS','PEST_TAXONOMY_SPECIES'),
          #keep the y in each
          by.y=c('Genus','Species'), all.y = TRUE)
    #determine which group it is by identifying the first entry in the identify.by column within the subset
  } else if (subset$identify.by[1] == "genus only") {
    # For this group, only the genus is identified in the high/low impact dataset.
    # Looking through the ARM records, when genus is identified, family is also identified
    #   see
       #view(df_diagnostic_results |> filter(!is.na(TAXONOMY_SPECIES)) |> count(TAXONOMY_FAMILY))
    merge(df_diagnostic_results, subset, by.x=c('PEST_TAXONOMY_FAMILY','PEST_TAXONOMY_GENUS'),
          by.y=c('Family','Genus'), all.y = TRUE)
  } else {
    # the only other group is genus and species, so default to comparing genus and species
    merge(df_diagnostic_results, subset, by.x=c('PEST_TAXONOMY_GENUS','PEST_TAXONOMY_SPECIES'),
          by.y=c('Genus','Species'), all.y=TRUE)
  }
})

# Combine the joined dataframes
# use dplyr::bind_rows instead of rbind because the matching columns rearrange the column order;
#   bind_rows will match column names instead of requiring the same order
HLI_final_df <- do.call(bind_rows, HLI_joined_list)

# #does it join correctly?
# df_diagnostic_results |> left_join(final_df)
# #is it matching genus and species correctly? do they have the same name?
#  df_diagnostic_results |> left_join(final_df) |> filter(!is.na(`Qualitative Impact Rating`)) |>
#    dplyr::select(TAXONOMY_GENUS,TAXONOMY_SPECIES,`Pest Scientific Name`)
# #has it duplicated anything?
# which(duplicated(df_diagnostic_results) == TRUE) #this code works correctly, as evidenced by changing TRUE to FALSE
# #everything appears to be working correctly

df_diagnostic_results = df_diagnostic_results |> left_join(HLI_final_df) |> dplyr::select(
  #all of complete.records columns
  colnames(df_diagnostic_results),
  #source of info, pest impact, and scientific name for completeness
  Source,
  `Qualitative Impact Rating`,
  `Pest Scientific Name`,
  HLI_score)



```

Diagnostic results table is now carrying columns that indicate if the pest was on the OPEP, FNW, or HLI list

## Consolidate that information

#### the hang up right now is that there is no state of the port of arrival

Need state or at least cnous/territory/hawaii of POE to check against the quarnatine status at each of those, to figure out if it is quarantine at the POE location.

DETERMIANTION_LOCAITON_STATE has DC for some Miami FL and MN entries; Not sure where determination would occur for HI and PR - it might make it easier if I could find that determination for interceptions in those locations always happen within them

To determine if the pest is quarantined at the port of entry, we will extract the port information from that given in df_diagnostic_results table, compare that with ARM Datamart's reference table, and determine the state of the PoE systematically

I left off here (6/2/25) - I just realized there's a QUARANTINE_STATUS column in the data already... Need to check that this is quarantine status at POE and not of mainland. If it is at POE, can just drop the following chunk

```{r identifying port of entry quarantine status}
# #Identify port number
# df_diagnostic_results <- df_diagnostic_results |> 
#   mutate(port_number = str_extract(PORT_OF_ENTRY_NAME, "(?<=\\().*?(?=\\))"))
# 
# #Identify the column that reports port number - just so it's easier to change if it ever changes on this curated table
# port_number_in_XwalkLocation_tbl <- 'ARM_CBPPortNumber'
# 
# #Get reference table
# port_state_lookup_tbl <- tbl(
#   db_conn,
#   sql("SELECT * FROM [PPQ_AQI_ARMDMV2].[dbo].[tbl_AQI_XwalkLocation]")
# ) |> dplyr::select(all_of(port_number_in_XwalkLocation_tbl),STATE_CD) |> 
#   collect()
# 
# df_diagnostic_results <- df_diagnostic_results |> 
#   left_join(port_state_lookup_tbl,
#             by = c('port_number' = port_number_in_XwalkLocation_tbl),
#             relationship = 'many-to-one'
#             )
# 
# poe_loc <- ifelse(
#   df_diagnostic_results$STATE_CD == 'PR', 'QUARANTINE_STATUS_PUERTO_RICO'
# )

# #check that QUARANTINE_STATUS is quarantine status at POE and not of mainland.
# mvw_diagnostic_results <- tbl(
#   db_conn,
#   sql("select * FROM [APHIS_Imports].[dbo].[mvw_Diagnostic_Results]")
# )
# 
# df_diagnostic_results <- mvw_diagnostic_results %>% 
#   filter(
#     INSPECTION_DATE >= begin_date,
#     INSPECTION_DATE <= end_date,
#     DETERMINATION_TYPE == 'Final ID',
#     SUBCATEGORY %in% c("CBP", "PIS", "SITC", "Non-PIS PPQ"),
#     !is.na(PEST_TAXONOMIC_NAME)
#   ) %>% 
#   collect()
# 
# #records where quarantine status in mainland is diff from PR or hawaii
# test <- df_diagnostic_results |> mutate(non_match = case_when(QUARANTINE_STATUS_MAINLAND != QUARANTINE_STATUS_HAWAII | QUARANTINE_STATUS_MAINLAND != QUARANTINE_STATUS_PUERTO_RICO ~ 1,
#                                                       .default = 0 )) |> 
#   dplyr::filter(non_match > 0)
# 
# #does quarantin estatus mathc hawaii or puerto rico?
# test2 <- test |> 
#   mutate(foreign_match = case_when(
#     QUARANTINE_STATUS == QUARANTINE_STATUS_HAWAII | QUARANTINE_STATUS == QUARANTINE_STATUS_PUERTO_RICO ~ 1,
#     .default = 0
#   )) |> 
#   dplyr::filter(foreign_match > 0)
# #a lot of these had no phytosanitary action AND they were not found in hawaii or puerto rico - but QUARANTINE_STATUS does not align with mainland quarantine status..
# 
# #where were the POEs?
# test3 <- df_diagnostic_results |> filter(QUARANTINE_STATUS != QUARANTINE_STATUS_MAINLAND) |> count(PORT_OF_ARRIVAL_NAME)
# # I see entries from PR and HI, so this is not following just QUARANTINE STATUS MAINLAND
# #not exactly sure what QUARANTINE_STATUS is reflecting, though...
# test4 <- df_diagnostic_results |> 
#   filter(QUARANTINE_STATUS != QUARANTINE_STATUS_MAINLAND) |> 
#   count(PORT_OF_ARRIVAL_NAME,QUARANTINE_RECOMMENDATION)
# #The majority of these are no phytosanitary action as well
# 
# #Think I just need to get up with DART and see what this column is about.

#The answer is you can't use this column.. Need to just go with quarantine status 
```

Left off here 6/2/25

```{r quantification of impact}
#case_when() requires sequential ranking. If something is evaluated and fits the case of a prior question, it will not be questioned again
df_diagnostic_results <- df_diagnostic_results |> mutate(pest.risk = case_when(
  #program pests are the most significant, so use their score first
  is.na(HLI_score) == FALSE ~ HLI_score,
  #FNWs and OPEPs are next
  is.na(FNW_score) == FALSE ~ FNW_score,
  #Check for OPEP score here
  is.na(Predicted_pest_impact_in_US) == FALSE ~ Predicted_pest_impact_in_US,
  #all others will be evaluated for their ARM quarantine status
  quarantine.status.at.port == 'Quarantine' ~ quarantine.pest.value,
  quarantine.status.at.port == 'Uncategorized' ~ uncategorized.pest.value,
  quarantine.status.at.port == 'Non-Quarantine' ~ nonquarantine.pest.value,
  #anything that doesn't fit these is probably an error or needs to be reviewed; return as 0
  .default = 0
))
```

```{r summary pest impact by commodity}
commodity_pest_risk = df_diagnostic_results |> 
  group_by(PORT_OF_ENTRY_NAME,INSPECTION_PATHWAY,COMMODITY,ORIGIN_NM) |> summarise(mean.pest.risk = mean(pest.risk) )

#all combinations should be unique at this point (without mean.pest.risk)
commodity_pest_risk |> dplyr::select(-mean.pest.risk)
commodity_pest_risk |> dplyr::select(-mean.pest.risk) |> unique()

pathway.assessment = commodity_pest_risk |> 
  left_join(ARM.num.records.at.port, by=c('CBP_PORT_NUMBER','PATHWAY','COMMODITY','ORIGIN_NM')) |> 
  left_join(ARM.avg.live.pests.per.infestation.at.port, by=c('CBP_PORT_NUMBER','PATHWAY','COMMODITY','ORIGIN_NM')) |> 
  left_join(ARM.richness.pests.at.port, by=c('CBP_PORT_NUMBER','PATHWAY','COMMODITY','ORIGIN_NM')) |> 
  left_join(ARM.richness.family.at.port, by=c('CBP_PORT_NUMBER','PATHWAY','COMMODITY','ORIGIN_NM')) |> 
  left_join(ARM.num.shipments.with.live.pest.at.port, by=c('CBP_PORT_NUMBER','PATHWAY','COMMODITY','ORIGIN_NM')) |> 
  arrange(desc(mean.pest.risk))
```

Now find out where the port profile gets commodity information.
