---
title: "Getting Started with AgCargoPestRisk"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with AgCargoPestRisk}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE  # Set to TRUE when package is fully functional
)
```

# Introduction

The `AgCargoPestRisk` package provides a comprehensive toolkit for assessing pest risks associated with agricultural cargo imports. This vignette will walk you through the basic usage patterns.

## Installation

```{r installation}
# Install from local source
devtools::install_local("path/to/AgCargoPestRisk")

# Or from GitHub (when available)
# devtools::install_github("USDA/AgCargoPestRisk")
```

## Quick Start: Complete Pipeline

The simplest way to run a complete pest risk assessment is using the `run_pest_risk_pipeline()` function:

```{r complete-pipeline}
library(AgCargoPestRisk)

results <- run_pest_risk_pipeline(
  begin_date = "2019-10-01",
  end_date = "2025-04-01",
  opep_directory = "path/to/OPEP",
  fnw_directory = "path/to/FNW",
  hli_directory = "path/to/HLI",
  output_dir = "outputs",
  export_results = TRUE
)

# Access results
pathway_assessment <- results$pathway_assessment
risk_summary <- results$risk_summary
```

### What the Pipeline Does

1. **Connects to database** and retrieves diagnostic results
2. **Loads pest impact data** from OPEP, FNW, and HLI sources
3. **Matches taxonomies** across different pest databases
4. **Calculates risk scores** based on hierarchy of impact assessments
5. **Summarizes by commodity and origin** for pathway assessment
6. **Exports results** to CSV and RDS formats

## Viewing Results

```{r view-results}
# Top 10 highest risk pathways
head(pathway_assessment, 10)

# Risk summary by port
print(risk_summary)

# Filter for specific port
miami_risks <- pathway_assessment %>%
  filter(grepl("Miami", PORT_OF_ENTRY_NAME))
```

## Customizing Parameters

You can adjust risk scoring parameters:

```{r custom-parameters}
results <- run_pest_risk_pipeline(
  begin_date = "2024-01-01",
  end_date = "2024-12-31",
  opep_directory = "path/to/OPEP",
  fnw_directory = "path/to/FNW",
  hli_directory = "path/to/HLI",
  # Custom risk values
  quarantine_pest_value = 0.6,      # Higher weight for quarantine pests
  uncategorized_pest_value = 0.3,
  nonquarantine_pest_value = 0.1,   # Small weight even for non-Q
  # Custom program pest scores
  caps_score = 1.0,
  ppq_program_score = 0.9,
  output_dir = "custom_outputs"
)
```

## Understanding Risk Scores

Risk scores are calculated using a hierarchical approach:

1. **Program Pests** (HLI): Highest priority (score: 0.3-1.0)
2. **FNW Assessments**: WRA-based invader probability (score: 0-1.0)
3. **OPEP Assessments**: Expert-derived impact scores (score: 0-1.0)
4. **Quarantine Status**: Default regulatory status
   - Quarantine: 0.5
   - Uncategorized: 0.25
   - Non-Quarantine: 0.0

The final score reflects the **likelihood and severity** of pest impact if introduced.

## Next Steps

- See `vignette("modular-usage")` for using individual functions
- See `vignette("advanced-customization")` for custom workflows
- Check `?run_pest_risk_pipeline` for detailed parameter documentation

## Common Use Cases

### Port-Specific Analysis

```{r port-analysis}
# Get all data, then filter
all_results <- run_pest_risk_pipeline(...)

port_assessment <- all_results$pathway_assessment %>%
  filter(PORT_OF_ENTRY_NAME == "Miami, FL") %>%
  arrange(desc(mean.pest.risk))
```

### Commodity-Specific Analysis

```{r commodity-analysis}
mango_risks <- all_results$pathway_assessment %>%
  filter(COMMODITY == "Mango") %>%
  arrange(desc(mean.pest.risk))
```

### Time Period Comparison

```{r time-comparison}
# Run for different periods
results_2023 <- run_pest_risk_pipeline(
  begin_date = "2023-01-01",
  end_date = "2023-12-31",
  ...
)

results_2024 <- run_pest_risk_pipeline(
  begin_date = "2024-01-01",
  end_date = "2024-12-31",
  ...
)

# Compare results
comparison <- bind_rows(
  results_2023$risk_summary %>% mutate(year = 2023),
  results_2024$risk_summary %>% mutate(year = 2024)
)
```
