---
title: "Modular Usage: Using Individual Functions"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Modular Usage: Using Individual Functions}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

# Modular Usage

While the complete pipeline is convenient, you may want to use individual functions for:

- **Custom workflows** that don't fit the standard pipeline
- **Integration** with existing analysis code
- **Debugging** specific steps
- **Performance** optimization by caching intermediate results

## Module A: Database Functions

### Connecting to the Database

```{r database-connection}
library(AgCargoPestRisk)

# Connect to PPQ database
conn <- connect_db()
```

### Retrieving Diagnostic Results

```{r get-diagnostics}
# Get diagnostic results for date range
diagnostic_data <- get_diagnostic_results(
  connection = conn,
  begin_date = "2024-01-01",
  end_date = "2024-12-31",
  limit = 50000  # Adjust as needed
)

# Get determination details
determination_data <- get_determinations(conn)

# Always close connection when done
DBI::dbDisconnect(conn)
```

### Processing Diagnostic Data

```{r process-diagnostics}
# Join and parse to final determinations
diagnostic_data <- diagnostic_data %>%
  left_join(determination_data, 
           by = intersect(colnames(diagnostic_data), 
                         colnames(determination_data))) %>%
  parse_final_determination()
```

## Module B: Data Processing

### Loading Pest Impact Data

```{r load-pest-data}
# Load OPEP data
opep_data <- load_opep_data("path/to/OPEP")

# Load FNW data with custom quarantine value
fnw_data <- load_fnw_data("path/to/FNW", quarantine_pest_value = 0.5)

# Load HLI data with custom scores
hli_data <- load_hli_data(
  "path/to/HLI",
  caps_score = 1.0,
  opep_score = 0.3,
  ppq_program_score = 1.0,
  select_agents_score = 1.0
)
```

## Module C: Risk Assessment

### Integrating Pest Assessments

You can integrate assessments individually or in combination:

```{r integrate-assessments}
# Option 1: Integrate only OPEP
diagnostic_data_opep <- integrate_opep_impacts(diagnostic_data, opep_data)

# Option 2: Integrate all three
diagnostic_data_full <- diagnostic_data %>%
  integrate_opep_impacts(opep_data) %>%
  integrate_fnw_impacts(fnw_data) %>%
  integrate_hli_impacts(hli_data)
```

### Calculating Risk Scores

```{r calculate-risk}
# Calculate final risk scores
diagnostic_data_scored <- calculate_pest_risk(
  diagnostic_data_full,
  quarantine_value = 0.5,
  uncategorized_value = 0.25,
  nonquarantine_value = 0
)
```

## Module D: Analysis Functions

### Summarizing Results

```{r summarize}
# Summarize by commodity-origin combinations
commodity_summary <- summarize_commodity_pest_risk(diagnostic_data_scored)

# Create pathway assessment
pathway_data <- create_pathway_assessment(commodity_summary)

# Generate risk statistics
port_summary <- generate_risk_summary(pathway_data, 
                                     group_vars = "PORT_OF_ENTRY_NAME")

commodity_summary <- generate_risk_summary(pathway_data, 
                                          group_vars = "COMMODITY")
```

### Exporting Results

```{r export}
# Export with timestamp
export_results(pathway_data, "pathway_assessment", "outputs")

# Export without timestamp (for automated pipelines)
export_results(pathway_data, "latest_assessment", "outputs", 
              include_timestamp = FALSE)
```

## Example: Custom Workflow for Specific Port

Here's a complete example of analyzing just one port with only OPEP data:

```{r custom-workflow}
library(AgCargoPestRisk)
library(dplyr)

# 1. Get data
conn <- connect_db()
diagnostic_data <- get_diagnostic_results(conn, "2024-01-01", "2024-12-31")
determ_data <- get_determinations(conn)
DBI::dbDisconnect(conn)

# 2. Process and filter for Miami
miami_data <- diagnostic_data %>%
  left_join(determ_data, by = intersect(colnames(.), colnames(determ_data))) %>%
  parse_final_determination() %>%
  filter(grepl("Miami", PORT_OF_ENTRY_NAME, ignore.case = TRUE))

# 3. Load only OPEP data (skip FNW and HLI)
opep_data <- load_opep_data("path/to/OPEP")

# 4. Integrate OPEP and calculate risk
miami_scored <- miami_data %>%
  integrate_opep_impacts(opep_data) %>%
  mutate(
    # Custom risk calculation
    risk_score = case_when(
      !is.na(Predicted_pest_impact_in_US) ~ Predicted_pest_impact_in_US,
      QUARANTINE_STATUS_CONUS == "Q" ~ 0.5,
      TRUE ~ 0
    )
  )

# 5. Summarize
miami_summary <- miami_scored %>%
  group_by(COMMODITY, ORIGIN_NM) %>%
  summarise(
    detections = n(),
    mean_risk = mean(risk_score, na.rm = TRUE),
    opep_pests = sum(!is.na(Predicted_pest_impact_in_US)),
    .groups = 'drop'
  ) %>%
  arrange(desc(mean_risk))

# 6. Export
export_results(miami_summary, "miami_opep_analysis", "outputs")
```

## Example: Adding Custom Risk Factors

You can extend the risk assessment with your own factors:

```{r custom-factors}
# Add custom risk adjustments
diagnostic_data_custom <- diagnostic_data_scored %>%
  mutate(
    # Increase risk for certain families
    family_risk_multiplier = case_when(
      PEST_TAXONOMY_FAMILY == "Tephritidae" ~ 1.2,  # Fruit flies
      PEST_TAXONOMY_FAMILY == "Curculionidae" ~ 1.1,  # Weevils
      TRUE ~ 1.0
    ),
    # Adjust based on time of year (hypothetical)
    seasonal_risk = case_when(
      month(INSPECTION_DATE) %in% c(6, 7, 8) ~ 1.1,  # Summer
      TRUE ~ 1.0
    ),
    # Calculate adjusted risk
    adjusted_risk = pest.risk * family_risk_multiplier * seasonal_risk
  )

# Summarize with custom risk
custom_summary <- diagnostic_data_custom %>%
  group_by(PORT_OF_ENTRY_NAME, COMMODITY, ORIGIN_NM) %>%
  summarise(
    mean_standard_risk = mean(pest.risk, na.rm = TRUE),
    mean_adjusted_risk = mean(adjusted_risk, na.rm = TRUE),
    risk_increase = mean_adjusted_risk - mean_standard_risk,
    .groups = 'drop'
  ) %>%
  arrange(desc(mean_adjusted_risk))
```

## Tips for Modular Usage

1. **Cache intermediate results**: Save outputs of expensive operations (database queries, data loading) as RDS files
2. **Use pipes**: Chain functions together for readable workflows
3. **Mix and match**: Combine package functions with your own custom code
4. **Start simple**: Begin with one module, then expand as needed
5. **Document dependencies**: Note which functions depend on outputs from others
