# Module C: Analysis and Summarization

This module creates summaries and pathway assessments from pest risk data.
Now simplified using the AgCargoPestRisk package functions.

## Overview
The purpose of this module is to take pest taxonomies with impact scores and summarize them.
This handles both:
1. Simple pest lists with impact scores (pest taxonomy summaries)
2. Complete datasets with shipment details (pathway and commodity summaries)

```{r load libraries, message=FALSE, warning=FALSE}
# Source package functions during development
source("../R/analysis_functions.R")
source("../R/utils.R")

library(dplyr)
library(readr)
library(ggplot2)
library(binom)
```

```{r load data from module b}
# Load pest risk data from Module B output
input_file <- "../outputs/manifest_example_module_b_output.rds"

if (file.exists(input_file)) {
  records <- readRDS(input_file)
  message("Loaded ", nrow(records), " records from Module B")
} else {
  # Fallback to CSV if RDS not available
  csv_file <- "../outputs/manifest_example_module_b_output.csv"
  if (file.exists(csv_file)) {
    records <- read_csv(csv_file)
    message("Loaded ", nrow(records), " records from Module B (CSV)")
  } else {
    stop("No Module B output found. Please run Module B first.")
  }
}
```

# Summarize pest impact rating sources

```{r summarize pest taxonomies}

# Summary of risk score distribution
impact_summary <- records %>%
  summarise(
    total_records = n(),
    mean_risk = mean(pest_impact, na.rm = TRUE),
    high_risk_count = sum(pest_impact > 0.5, na.rm = TRUE),
    medium_risk_count = sum(pest_impact >= 0.25 & pest_impact <= 0.5, na.rm = TRUE),
    low_risk_count = sum(pest_impact > 0 & pest_impact < 0.25, na.rm = TRUE),
    zero_risk_count = sum(pest_impact == 0, na.rm = TRUE),
    quarantined = sum(quarantine_status_at_port == "Quarantine", na.rm = TRUE)
  )

print("Impact Score Summary:")
print(impact_summary)
View(impact_summary)
```

# Create a parent summary table
Can build in utility for this, such as an web-based interface that allows drill down charts

```{r pest impact by pathways}
parent_summary_table <- records |> 
  dplyr::select(
    INSPECTION_DATETIME,

    COMMODITY_COMMON_NAME,
    COUNTRY_OF_ORIGIN_NAME,
    PATHWAY,
    REGION,
    LOCATION_STATE_CODE,
    CBP_PORT_NUMBER,

    HOST_PROXIMITY,
    QUANTITY,

    PEST_TAXON_SIMPLE_NAME,
    NUMBER_ALIVE_ADULTS,
    NUMBER_ALIVE_PUPAE,
    NUMBER_ALIVE_CYSTS,
    NUMBER_ALIVE_EGGS,
    NUMBER_ALIVE_IMMATURE,

    pest_impact
  )

```

# Create pathway summary
```{r pathway summary}
port_pathway_summary <- parent_summary_table |>
  group_by(
    COMMODITY_COMMON_NAME,
    COUNTRY_OF_ORIGIN_NAME,
    PATHWAY,
    REGION,
    LOCATION_STATE_CODE,
    CBP_PORT_NUMBER
  ) %>%
  summarise(
    total_detections = n(),
    mean_risk = mean(pest_impact, na.rm = TRUE),
    high_risk_detections = sum(pest_impact > 0.5, na.rm = TRUE),
    medium_risk_detections = sum(pest_impact >= 0.25 & pest_impact <= 0.5, na.rm = TRUE),
    low_risk_detections = sum(pest_impact > 0 & pest_impact < 0.25, na.rm = TRUE),
    zero_risk_detections = sum(pest_impact == 0, na.rm = TRUE),
    .groups = 'drop',
    avg_n_life_stages = (sum(
      !is.na(NUMBER_ALIVE_ADULTS)) + 
      sum(!is.na(NUMBER_ALIVE_PUPAE)) + 
      sum(!is.na(NUMBER_ALIVE_EGGS)) + 
      sum(!is.na(NUMBER_ALIVE_IMMATURE)) + 
      sum(!is.na(NUMBER_ALIVE_CYSTS))) / n()
  ) %>%
  arrange(desc(mean_risk))

#calculate probability of a high risk pest
port_pathway_summary <- port_pathway_summary |>
  mutate(
    #calculate basic proportion
    prob_high_risk_pest = high_risk_detections / total_detections
  ) |>
  # add confidence intervals
  rowwise() %>%
  mutate(ci = list(
    binom.confint(
      high_risk_detections, 
      total_detections, 
      methods = "wilson", 
      conf.level = 0.95)
      ), 
      ci_lower_95 = ci$lower, 
      ci_upper_95 = ci$upper ) |>
       ungroup() |> 
       select(-ci) |>
       arrange(desc(prob_high_risk_pest))




print(head(port_pathway_summary, 15))
View(port_pathway_summary)

```