# Module C: Analysis and Summarization

This module creates summaries and pathway assessments from pest risk data.
Now simplified using the AgCargoPestRisk package functions.

## Overview
The purpose of this module is to take pest taxonomies with impact scores and summarize them.
This handles both:
1. Simple pest lists with impact scores (pest taxonomy summaries)
2. Complete datasets with shipment details (pathway and commodity summaries)

```{r load libraries, message=FALSE, warning=FALSE}
# Source package functions during development
source("../R/analysis_functions.R")
source("../R/utils.R")

library(dplyr)
library(readr)
library(ggplot2)
library(binom)
```

```{r load data from module b}
# Load pest risk data from Module B output
input_file <- "../outputs/module_b_output.rds"

if (file.exists(input_file)) {
  records <- readRDS(input_file)
  message("Loaded ", nrow(records), " records from Module B")
} else {
  # Fallback to CSV if RDS not available
  csv_file <- "../outputs/module_b_output.csv"
  if (file.exists(csv_file)) {
    records <- read_csv(csv_file)
    message("Loaded ", nrow(records), " records from Module B (CSV)")
  } else {
    stop("No Module B output found. Please run Module B first.")
  }
}
```

# Summarize pest impact rating sources

```{r summarize pest taxonomies}

# Summary of risk score distribution
impact_summary <- records %>%
  summarise(
    total_records = n(),
    mean_risk = mean(pest_impact, na.rm = TRUE),
    high_risk_count = sum(pest_impact > 0.5, na.rm = TRUE),
    medium_risk_count = sum(pest_impact >= 0.25 & pest_impact <= 0.5, na.rm = TRUE),
    low_risk_count = sum(pest_impact > 0 & pest_impact < 0.25, na.rm = TRUE),
    zero_risk_count = sum(pest_impact == 0, na.rm = TRUE),
    quarantined = sum(quarantine_status_at_port == "Quarantine", na.rm = TRUE)
  )

print("Impact Score Summary:")
print(impact_summary)
```

# Create a parent summary table
Can build in utility for this, such as an web-based interface that allows drill down charts

```{r pest impact by pathways}
parent_summary_table <- records |> 
  dplyr::select(
    INSPECTION_DATETIME,

    COMMODITY_COMMON_NAME,
    COUNTRY_OF_ORIGIN_NAME,
    PATHWAY,
    REGION,
    LOCATION_STATE_CODE,
    CBP_PORT_NUMBER,

    HOST_PROXIMITY,
    QUANTITY,

    PEST_TAXON_SIMPLE_NAME,
    NUMBER_ALIVE_ADULTS,
    NUMBER_ALIVE_PUPAE,
    NUMBER_ALIVE_CYSTS,
    NUMBER_ALIVE_EGGS,
    NUMBER_ALIVE_IMMATURE,

    pest_impact
  )

```

# Create pathway summary
```{r pathway summary}
port_pathway_summary <- parent_summary_table |>
  group_by(
    COMMODITY_COMMON_NAME,
    COUNTRY_OF_ORIGIN_NAME,
    PATHWAY,
    REGION,
    LOCATION_STATE_CODE,
    CBP_PORT_NUMBER
  ) %>%
  summarise(
    total_detections = n(),
    mean_risk = mean(pest_impact, na.rm = TRUE),
    high_risk_detections = sum(pest_impact > 0.5, na.rm = TRUE),
    medium_risk_detections = sum(pest_impact >= 0.25 & pest_impact <= 0.5, na.rm = TRUE),
    low_risk_detections = sum(pest_impact > 0 & pest_impact < 0.25, na.rm = TRUE),
    zero_risk_detections = sum(pest_impact == 0, na.rm = TRUE),
    .groups = 'drop',
    avg_n_life_stages = (sum(
      !is.na(NUMBER_ALIVE_ADULTS)) + 
      sum(!is.na(NUMBER_ALIVE_PUPAE)) + 
      sum(!is.na(NUMBER_ALIVE_EGGS)) + 
      sum(!is.na(NUMBER_ALIVE_IMMATURE)) + 
      sum(!is.na(NUMBER_ALIVE_CYSTS))) / n()
  ) %>%
  arrange(desc(mean_risk))

#calculate probability of a high risk pest
port_pathway_summary <- port_pathway_summary |>
  mutate(
    #calculate basic proportion
    prob_high_risk_pest = high_risk_detections / total_detections
  ) |>
  # add confidence intervals
  rowwise() %>%
  mutate(ci = list(
    binom.confint(
      high_risk_detections, 
      total_detections, 
      methods = "wilson", 
      conf.level = 0.95)
      ), 
      ci_lower_95 = ci$lower, 
      ci_upper_95 = ci$upper ) |>
       ungroup() |> 
       select(-ci) |>
       arrange(desc(prob_high_risk_pest))




print(head(port_pathway_summary, 15))
View(port_pathway_summary)

```


```{r summarize by commodity and pathway}
# Create commodity-pathway summaries using package function
commodity_pest_risk <- summarize_commodity_pest_risk(df_diagnostic_results)

# Create pathway assessment using package function
pathway_assessment <- create_pathway_assessment(commodity_pest_risk)

message("Created pathway assessment with", nrow(pathway_assessment), "commodity-origin combinations")
```

```{r generate summary statistics}
# Generate risk summary by port using package function
port_summary <- generate_risk_summary(pathway_assessment, group_vars = "PORT_OF_ENTRY_NAME")

# Generate risk summary by commodity
commodity_summary <- generate_risk_summary(pathway_assessment, group_vars = "COMMODITY")

# Generate risk summary by origin country
origin_summary <- generate_risk_summary(pathway_assessment, group_vars = "ORIGIN_NM")

# Overall summary statistics
overall_summary <- df_diagnostic_results %>%
  summarise(
    total_detections = n(),
    unique_pests = n_distinct(paste(PEST_TAXONOMY_GENUS, PEST_TAXONOMY_SPECIES)),
    unique_ports = n_distinct(PORT_OF_ENTRY_NAME),
    unique_commodities = n_distinct(COMMODITY),
    unique_origins = n_distinct(ORIGIN_NM),
    mean_risk_score = mean(pest.risk, na.rm = TRUE),
    high_risk_detections = sum(pest.risk >= 0.5, na.rm = TRUE),
    high_risk_percentage = round(100 * sum(pest.risk >= 0.5, na.rm = TRUE) / n(), 2)
  )

print("Overall Summary:")
print(overall_summary)
```

```{r create visualizations}
# Risk score distribution
risk_dist_plot <- ggplot(df_diagnostic_results, aes(x = pest.risk)) +
  geom_histogram(bins = 30, fill = "steelblue", alpha = 0.7) +
  labs(
    title = "Distribution of Pest Risk Scores",
    x = "Risk Score",
    y = "Count"
  ) +
  theme_minimal()

# Top risk commodity-origin combinations
top_risks_plot <- pathway_assessment %>%
  head(20) %>%
  mutate(combo = paste(COMMODITY, "-", ORIGIN_NM)) %>%
  ggplot(aes(x = reorder(combo, mean.pest.risk), y = mean.pest.risk)) +
  geom_col(fill = "darkred", alpha = 0.7) +
  coord_flip() +
  labs(
    title = "Top 20 Highest Risk Commodity-Origin Combinations",
    x = "Commodity - Origin",
    y = "Mean Risk Score"
  ) +
  theme_minimal()

# Risk by port
port_risk_plot <- port_summary %>%
  ggplot(aes(x = reorder(PORT_OF_ENTRY_NAME, mean_risk_score), y = mean_risk_score)) +
  geom_col(fill = "darkgreen", alpha = 0.7) +
  coord_flip() +
  labs(
    title = "Mean Risk Score by Port",
    x = "Port of Entry",
    y = "Mean Risk Score"
  ) +
  theme_minimal()

print(risk_dist_plot)
print(top_risks_plot)
print(port_risk_plot)
```

```{r export module c results}
# Export all results using package function
output_dir <- "../outputs"

export_results(pathway_assessment, "pathway_assessment", output_dir)
export_results(pest_taxonomy_summary, "pest_taxonomy_summary", output_dir)
export_results(port_summary, "port_risk_summary", output_dir)
export_results(commodity_summary, "commodity_risk_summary", output_dir)
export_results(origin_summary, "origin_risk_summary", output_dir)

# Save plots
ggsave(file.path(output_dir, "risk_distribution.png"), risk_dist_plot, width = 10, height = 6)
ggsave(file.path(output_dir, "top_risk_combinations.png"), top_risks_plot, width = 12, height = 8)
ggsave(file.path(output_dir, "port_risk_summary.png"), port_risk_plot, width = 10, height = 6)

message("Module C analysis complete!")
message("All outputs saved to:", output_dir)
```