# Module A: Data Retrieval and Processing

This module handles database connections, data retrieval, and initial processing.
Now simplified using the AgCargoPestRisk package functions.

## Overview
- Load parameters
- Retrieve inspections/commodities data based on these parameters
- Retrieve connected interceptions based on these parameters
- Process data to final determinations
- Identify quarantine status of pests identified
- Output pest taxonomies, location, region, state, port, commodities, start date, end date

```{r load libraries, message=FALSE, warning=FALSE}
# Option 1: Source functions during development
# Use here::here() to get paths relative to project root
source(here::here("R", "database_functions.R"))
source(here::here("R", "data_processing.R"))
source(here::here("R", "utils.R"))

# Option 2: Load as package (after running devtools::load_all() or devtools::install())
# library(AgCargoPestRisk)

library(dplyr)
library(readr)
library(DBI)
library(here)
```

# Load parameters
- this needs to be edited to use a list from params

## Port or Region approach
A port number or region name can be used by an ag specialist to take a look at the shipments coming into that port or region

## Manifest approach
An alternative use would be to apply a manifest. `pathway_manifest` is a placeholder for a system where one could connect a set of commodity names and origin countries from shipments that are expected to come in.

```{r load parameters}
# Load parameters from YAML or define here
begin_date = "2019-10-01"
end_date = "2025-04-01"

# Additional parameters for filtering
port_regions <- NULL  # specify if needed
state <- NULL   # Not currently supported; build into get_f280_data() function
port_numbers <- c("2704")    # specify if needed
commodities <- NULL  # specify if needed
origin_country <- NULL # specify if needed
pathway_manifest <- NULL # specify if using a manifest to filter
```

# First step is to identify what pathways exist at your port of region of interest
 - This reduces the amount of searching that is done against ARM
 - This should be skipped if supplying a pathway_manifest
## Check into time range issues
I'm having trouble when i shrink the time range from 2019 to 2023 and then re-run; but it works when the time range is at 2019. Not sure where it is coming from right now. Something about how it is being translated to SQL code, maybe?

```{r retrieve F280 data}
# get the information about the pathways that exist and the number of shipments in that pathway
db_conn <- connect_db()
pathways <- get_f280_data(db_conn,
              date_start = begin_date,
              date_end = end_date,
              #if necessary, specifiy port numbers
              port_numbers = port_numbers,
              #if necessary, specify region
              port_regions = port_regions,
              #limit = 20000 # limit can be used when in development in order to reduce the download time
              ) |> 
              process_num_ship()
```
remove the following if no more utility:
```{r}

# # Use this information in the next steps in order to reduce the amount of information retrieved from ARM
# commodities <- if(
#   !exists("commodities")) {
#     pathways |> distinct(COMMODITY) |> pull(COMMODITY)
#   } else {commodities}

# origin_country <- if(
#   !exists("origin_country")) {
#     pathways |> distinct(ORIGIN_NM) |> pull(ORIGIN_NM)
#   } else {origin_country}

# port_numbers <- if(
#   !exists("port_numbers")) {
#     pathways |> distinct(CBP_PORT_NUMBER) |> pull(CBP_PORT_NUMBER)
#   } else {port_numbers}

# port_regions <- if(
#   !exists("port_regions")) {
#     pathways |> distinct(REGION) |> pull(REGION)
#   } else {port_regions}

```


# Gather inspections and commodity data from ARM
 Connects the following shipment information:
  - inspection information
  - commodity information
  - port information

## Check into location issues
I'm seeing records for all kinds of locations
 - check if I'm missing a layer of filtering; perhaps by port?

```{r}
# commodity cannot be filtered by date time or country of origin; need inspection table to do that
# get_inspection_id will by fitlered by date time and provide a list of inspection IDs to filter for commodities; other filtering will occur for commodities
# NOTE TO SELF: THIS FUNCTION SHOULD FOCUS ON FILTERING DATETIME AND COUNTRY

system.time(inspection_df <- get_inspections(connection = db_conn,
                          date_start = begin_date,
                          date_end = end_date,
                          origin_country = origin_country)
)

inspection_ids <- get_inspection_ids(inspection_df)

# Gather commodity info
  # Where 'commodities' == commodities identified in f280, or elsewhere

system.time(commodity_df <- get_commodity(connection = db_conn,
                            commodities = commodities))

# Gather the location info from ARM, which allows connecting the port number and region, etc. to the inspection and commodity table
system.time(location_df <- get_location(
  connection = db_conn,
  date_start = begin_date,
  date_end = end_date
))


# Merge inspection, commodity, and location data
shipment_records <- merge_records(
  commodity_df = commodity_df,
  inspection_df = inspection_df,
  location_df = location_df
)

# Filter to only keep commodity-origin combinations that exist in pathways (F280 data)
# This ensures we only analyze shipments for pathways that were actually reported in F280
if(exists("pathways") & is.null(pathway_manifest)) {
shipment_records <- shipment_records %>%
  semi_join(
    pathways %>% 
      distinct(COMMODITY, ORIGIN_NM, PORT_CD, REGION),
    by = c("COMMODITY_COMMON_NAME" = "COMMODITY", 
           "COUNTRY_OF_ORIGIN_NAME" = "ORIGIN_NM",
           "CBP_PORT_NUMBER" = "PORT_CD",
           "REGION" = "REGION"
           )
  )
} else if (!is.null("pathway_manifest")) {
  shipment_records <- shipment_records %>%
    semi_join(
      pathway_manifest,
      by = c("COMMODITY_COMMON_NAME" = "COMMODITY",
             "COUNTRY_OF_ORIGIN_NAME" = "ORIGIN_NM",
             "CBP_PORT_NUMBER" = "PORT_CD"
      )
    )
}

message("After filtering : ", nrow(shipment_records), " shipment records remain")


#If it was engaged by a pathway_manifest (combination of both commodities and their origin country) instead of a specific port or region (e.g., applying to a manifest), then it won't be filtered again at this stage because it doesn't need to be
```



# Deprecated: SKIP
## a previous method the FO's [APHIS_Imports].[dbo].[mvw_Diagnostic_Results] table

FO's materialized view table is much easier to use, however, I have concerns about this table. It doesn't match SQL queries to ARM that we put in directly, and it has to be ETL'd (extract transform and load) by a data engineer. I don't know if this is an automatic process but the shutdown has given me some pause on relying on it.
  - It would be safer to take the more complex route of building our output from SQL queries to ARM DataMart tables directly.
  - If it were to be implemented in this enhanced approach, I would have to double check that this table also carries commodity records that did not find a pest
  
```{r retrieve inspections/commodities data, eval=FALSE}

db_conn <- connect_db()

# Retrieve diagnostic results using package function
df_diagnostic_results <- get_diagnostic_results_FO_method(
  connection = db_conn,
  begin_date = params$begin_date,
  end_date = params$end_date,
  limit = 10000
)

message(paste("Retrieved", nrow(df_diagnostic_results), "diagnostic records"))
```



# Gather the pest diagnostic information
Connects pest information (diagnostics) that are found on the shipments:
  - diagnostic request from a pest that was found
  - diagnostic determination from a requested submission (i.e., an identification tag)
    - diagnostic failures (e.g., molecular analysis failed)
  - enumeration of the alive, dead, and life stages that were associated with each diagnostic request


```{r retrieve connected interceptions and determinations}
#retrieve the diagnostic requests associated with our commodity IDs that we've identified
diagnostic_request_df <- get_diagnostic_requests(
  connection = db_conn,
  shipment_records = shipment_records
)

# Get determination details using package function
diag_determ_records <- get_determinations(
  db_conn,
  diagnostic_request_data = diagnostic_request_df
)

failed_determinations <- get_diag_determ_not_possible(
  connection = db_conn,
  diagnostic_determination_data = diag_determ_records
)

#retrieve the enumeration of the life stages and dead vs alive for each pest
dead_alive_df <- get_dead_alive(
  connection = db_conn,
  diagnostic_request_data = diagnostic_request_df
)

#retrieve the pest taxonomy information since the final determination doesn't actually give a name; it is connected with PEST_TAXON_ID
pest_taxonomy_df <- get_pest_taxonomy(
  connection = db_conn,
  pest_diagnostics = diag_determ_records
)

pest_diagnostics <- gather_diagnostics(
  diagnostic_requests = diagnostic_request_df,
  diagnostic_determinations = diag_determ_records,
  failed_determinations = failed_determinations,
  enumerated_life_stages = dead_alive_df,
  pest_taxonomy = pest_taxonomy_df
)

message(paste("Retrieved", nrow(pest_diagnostics), "pest diagnostic records"))
```

## Arrive at a single, final determination
Interceptions of various life stages or liveliness may lead to multiple diagnostic requests. This function reduces duplication that is created through this and several other processes.
```{r process data to final determinations}
# Process to get final determinations using package function
pest_diagnostics <- parse_final_determination(pest_diagnostics)


message(paste("After processing:", nrow(pest_diagnostics), "final determination records"))
```

# Merge shipment info with diagnostic results
 ## This chunk can still be converted into a function and stored elsewhere for clearer code here
```{r combine shipment and diagnostics}
# Merge pest diagnostics with shipment records to get full context
shipment_records <- shipment_records %>%
  left_join(pest_diagnostics, by = "COMMODITY_ID")

message(paste("After merging diagnostics:", nrow(shipment_records), "shipment records with diagnostics"))

# Collapse multiple life stage records for same commodity-pest combination
# Sum all NUMBER_* columns (life stages) for each unique commodity-pest pair
shipment_records_collapsed <- shipment_records %>%
  group_by(COMMODITY_ID, PEST_TAXON_SIMPLE_NAME) %>%
  summarise(
    # Sum all life stage counts
    across(starts_with('NUMBER_'), ~sum(.x, na.rm = TRUE)),
    # Keep first value for all other columns (not the grouping vars or NUMBER_ cols)
    across(!starts_with('NUMBER_') & !any_of(c("COMMODITY_ID", "PEST_TAXON_SIMPLE_NAME")), ~first(.x)),
    .groups = 'drop'
  )

message(paste("After collapsing life stages:", nrow(shipment_records) - nrow(shipment_records_collapsed), "duplicate pest identifications were removed"))

# Use the collapsed version for downstream analysis
shipment_records <- shipment_records_collapsed
rm(shipment_records_collapsed)
print(head(shipment_records))
```

# Save Module A output for further processing
At this point we have a list of interceptions, commodities, and pest information bundled up, with a focus on origin country and commodity but also tracking the port information.
 - Other use cases may extract this information and use it elsewhere.
 - We will pass it to Module B.

```{r export module A results}
# Save results for Module B
output_dir <- "../outputs"
if (!dir.exists(output_dir)) {
  dir.create(output_dir, recursive = TRUE)
}

# Export as CSV for universal compatibility
write_csv(shipment_records, file.path(output_dir, "module_a_output.csv"))

# Export as RDS for R-specific use (faster loading)
saveRDS(shipment_records, file.path(output_dir, "module_a_output.rds"))

message("Module A output saved to:", file.path(output_dir, "module_a_output.csv"))
message("Dataset contains ", nrow(shipment_records), " records with ", ncol(shipment_records), " columns")
```